Reanalysis Step 7: Sample Comparison Table

### 2025-09-04 reanalysis-7-missingness_patterns.qmd
data comparison

## Load Data from Three Sources

@@

# 2. Load complete cases from reanalysis (reanalysis-1)
complete_cases_file <- file.path(reanalysis_data_dir, "df_complete_original.rds")
if (!file.exists(complete_cases_file)) {
  stop("Complete cases data not found. Run reanalysis-1-data_preparation.qmd first.")
}
complete_cases_data <- readRDS(complete_cases_file)

log_message(paste0("Complete cases data: ", nrow(complete_cases_data), " observations"))

@@

```{r prepare-variables}
# Define variables for Table 1
vars_continuous <- c("age", "baseline_VA_logMAR", "CCI", "DCSI")
vars_categorical <- c("gender_cat", "race_ethnic_cat", "insurance_cat", "age_cat",
                      "dr_severity", "ever_lapse_binary", 
                      "glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat",
                      "catsurg_before_hitplus_cat",
                      "anti_VEGF", "PRP_flag", "focal_laser_flag", "other_inject",
                      "outcome_va_vi_binary")

all_vars <- c(vars_continuous, vars_categorical)

@@

# Process each dataset
process_dataset <- function(data, name) {
  # Continuous variables
  cont_summaries <- do.call(rbind, lapply(vars_continuous, function(v) {
    if (v %in% names(data)) {
      summarize_continuous(data, v)
    } else {
      data.frame(variable = v, mean = NA, sd = NA, median = NA, 
                 q25 = NA, q75 = NA, n_missing = NA, pct_missing = NA)
    }
  }))

  # Categorical variables
  cat_summaries <- do.call(rbind, lapply(vars_categorical, function(v) {
    if (v %in% names(data)) {
      summarize_categorical(data, v)
    } else {
      data.frame(variable = v, n = NA, pct = NA, n_missing = NA, pct_missing = NA)
    }
  }))

@@

```{r create-table}
# Process original data
original_summary <- process_dataset(original_complete, "Original")

# Process complete cases
complete_summary <- process_dataset(complete_cases_data, "Complete Cases")

@@

```{r display-table}
# Create formatted table for continuous variables
format_continuous_table <- function(original, complete, imputed) {
  vars <- unique(c(original$variable, complete$variable, imputed$variable))

  table_df <- data.frame(
    Variable = vars,
    Original = NA,
    Complete_Cases = NA,
    Imputed = NA,
    stringsAsFactors = FALSE
  )

@@

    # Complete cases
    comp_row <- complete[complete$variable == v, ]
    if (nrow(comp_row) > 0 && !is.na(comp_row$mean)) {
      table_df$Complete_Cases[i] <- sprintf("%.2f (%.2f)", comp_row$mean, comp_row$sd)
    }

@@

# Create formatted table for categorical variables
format_categorical_table <- function(original, complete, imputed) {
  vars <- unique(c(original$variable, complete$variable, imputed$variable))

  table_df <- data.frame(
    Variable = vars,
    Original = NA,
    Complete_Cases = NA, 
    Imputed = NA,
    stringsAsFactors = FALSE
  )

@@

    # Complete cases
    comp_row <- complete[complete$variable == v, ]
    if (nrow(comp_row) > 0 && !is.na(comp_row$n)) {
      table_df$Complete_Cases[i] <- sprintf("%.0f (%.1f%%)", comp_row$n, comp_row$pct)
    }

@@

# Create the tables
continuous_table <- format_continuous_table(
  original_summary$continuous,
  complete_summary$continuous,
  imputed_summary$continuous
)

categorical_table <- format_categorical_table(
  original_summary$categorical,
  complete_summary$categorical,
  imputed_summary$categorical
)

# Display continuous variables
kable(continuous_table, 
      caption = "Continuous Variables: Mean (SD)",
      col.names = c("Variable", "Original Data", "Complete Cases", "Imputed (Average)"))

# Display categorical variables
kable(categorical_table,
      caption = "Categorical Variables: N (%)",
      col.names = c("Variable", "Original Data", "Complete Cases", "Imputed (Average)"))

# Sample sizes
sample_sizes <- data.frame(
  Dataset = c("Original (after exclusions)", "Complete Cases", "Imputed (each)"),
  N = c(nrow(original_complete), nrow(complete_cases_data), nrow(imputed_datasets[[1]]))
)

>>>>

## Load Data from Four Sources

@@

# 2. Load new data with cohort assignments (from reanalysis-1)
cohort_file <- file.path(reanalysis_data_dir, "df_with_cohort.rds")
if (!file.exists(cohort_file)) {
  stop("Cohort data not found. Run reanalysis-1-data_preparation.qmd first.")
}
cohort_data <- readRDS(cohort_file)

# Split into submitted and excluded
submitted_data <- cohort_data[cohort_data$cohort_id == "submitted", ]
excluded_data <- cohort_data[cohort_data$cohort_id == "excluded", ]

log_message(paste0("Submitted (included) data: ", nrow(submitted_data), " observations"))
log_message(paste0("Excluded data: ", nrow(excluded_data), " observations"))

@@

```{r prepare-variables}
# Define variables for Table 1
# Note: Original data doesn't have dr_severity
vars_continuous <- c("age", "baseline_VA_logMAR", "CCI", "DCSI")

# Variables in original data (with proper capitalization)
vars_categorical_original <- c("gender_cat", "race_ethnic_cat", "insurance_cat", "age_cat",
                               "ever_lapse_binary", 
                               "Glaucoma_bef_hitplus_cat", "Otherretina_bef_hitplus_cat",
                               "catsurg_before_hitplus_cat",
                               "anti_VEGF", "PRP_flag", "focal_laser_flag", "other_inject",
                               "outcome_va_vi_binary")

# Additional variables in new data only
vars_categorical_new_only <- c("dr_severity")

# All categorical variables for new data
vars_categorical_new <- c("gender_cat", "race_ethnic_cat", "insurance_cat", "age_cat",
                          "dr_severity", "ever_lapse_binary", 
                          "Glaucoma_bef_hitplus_cat", "Otherretina_bef_hitplus_cat",
                          "catsurg_before_hitplus_cat",
                          "anti_VEGF", "PRP_flag", "focal_laser_flag", "other_inject",
                          "outcome_va_vi_binary")

@@

# Process each dataset
process_dataset <- function(data, name, categorical_vars = NULL) {
  # Use provided categorical vars or default to new vars
  if (is.null(categorical_vars)) {
    categorical_vars <- vars_categorical_new
  }

  # Continuous variables
  cont_summaries <- do.call(rbind, lapply(vars_continuous, function(v) {
    if (v %in% names(data)) {
      summarize_continuous(data, v)
    } else {
      data.frame(variable = v, mean = NA, sd = NA, median = NA, 
                 q25 = NA, q75 = NA, n_missing = NA, pct_missing = NA)
    }
  }))

  # Categorical variables
  cat_summaries <- do.call(rbind, lapply(categorical_vars, function(v) {
    if (v %in% names(data)) {
      summarize_categorical(data, v)
    } else {
      data.frame(variable = v, n = NA, pct = NA, n_missing = NA, pct_missing = NA)
    }
  }))

@@

```{r create-table}
# Process original data (with original variable names)
original_summary <- process_dataset(original_complete, "Original", vars_categorical_original)

# Process submitted (included) data
submitted_summary <- process_dataset(submitted_data, "Submitted", vars_categorical_new)

# Process excluded data
excluded_summary <- process_dataset(excluded_data, "Excluded", vars_categorical_new)

@@

```{r display-table}
# Create formatted table for continuous variables
format_continuous_table <- function(original, submitted, excluded, imputed) {
  vars <- unique(c(original$variable, submitted$variable, excluded$variable, imputed$variable))

  table_df <- data.frame(
    Variable = vars,
    Original = NA,
    Submitted = NA,
    Excluded = NA,
    Imputed = NA,
    stringsAsFactors = FALSE
  )

@@

    # Submitted
    sub_row <- submitted[submitted$variable == v, ]
    if (nrow(sub_row) > 0 && !is.na(sub_row$mean)) {
      table_df$Submitted[i] <- sprintf("%.2f (%.2f)", sub_row$mean, sub_row$sd)
    }

    # Excluded
    exc_row <- excluded[excluded$variable == v, ]
    if (nrow(exc_row) > 0 && !is.na(exc_row$mean)) {
      table_df$Excluded[i] <- sprintf("%.2f (%.2f)", exc_row$mean, exc_row$sd)
    }

@@

# Create formatted table for categorical variables
format_categorical_table <- function(original, submitted, excluded, imputed) {
  vars <- unique(c(original$variable, submitted$variable, excluded$variable, imputed$variable))

  table_df <- data.frame(
    Variable = vars,
    Original = NA,
    Submitted = NA,
    Excluded = NA,
    Imputed = NA,
    stringsAsFactors = FALSE
  )

@@

    # Submitted
    sub_row <- submitted[submitted$variable == v, ]
    if (nrow(sub_row) > 0 && !is.na(sub_row$n)) {
      table_df$Submitted[i] <- sprintf("%.0f (%.1f%%)", sub_row$n, sub_row$pct)
    }

    # Excluded
    exc_row <- excluded[excluded$variable == v, ]
    if (nrow(exc_row) > 0 && !is.na(exc_row$n)) {
      table_df$Excluded[i] <- sprintf("%.0f (%.1f%%)", exc_row$n, exc_row$pct)
    }

@@

# Create the tables
continuous_table <- format_continuous_table(
  original_summary$continuous,
  submitted_summary$continuous,
  excluded_summary$continuous,
  imputed_summary$continuous
)

categorical_table <- format_categorical_table(
  original_summary$categorical,
  submitted_summary$categorical,
  excluded_summary$categorical,
  imputed_summary$categorical
)

# Display continuous variables
kable(continuous_table, 
      caption = "Continuous Variables: Mean (SD)",
      col.names = c("Variable", "Original", "Submitted", "Excluded", "Imputed (Average)"))

# Display categorical variables
kable(categorical_table,
      caption = "Categorical Variables: N (%)",
      col.names = c("Variable", "Original", "Submitted", "Excluded", "Imputed (Average)"))

# Sample sizes
sample_sizes <- data.frame(
  Dataset = c("Original (after exclusions)", "Submitted (included)", "Excluded", "Imputed (each)"),
  N = c(nrow(original_complete), nrow(submitted_data), nrow(excluded_data), nrow(imputed_datasets[[1]]))
)
